---
title: "Atomic Vectors"
subtitle: "PFE I: Sessions IV & V"
author: "Prof. Dr. Juanjo Manjarín"
# date: "2 August 2019"
output:
  html_document:
    toc: true
    theme: cerulean
  html_notebook: default
  pdf_document: 
    toc: true
    theme: cerulean
---

## Introduction

One of the main ideas we have to keep in mind when we are coding in R is the following:

<p style="text-align: center";> 
"Every number in R is a double-precision floating-point vector"
</p>

This has two main implications:

  * **First**, that R does not work with integer numbers by default. To understand what this means let's take, for example, the number 2. If we look for its type:
 
```{r}
typeof(2)
```

  then, even in basic operations, we may find strange results: we may expect that if we do $(\sqrt{2})^2-2$ we will find $0$ as result, however
  
```{r}
sqrt(2)^2-2
```

  we obtain a very close to zero result, but not exactly. This can be a source of problems in certain computations that may require precision. See that the result is true even if

```{r}
sqrt(2)^2
```

If we may want to declare an integer number we have to use the L-flag by it

```{r}
typeof(2L)
```

then anytime we see a variable in the Global Environment with this flag by it, we must understand it as an integer value.

  * **Second**, that the native working procedure in R is that of vectorization. The meaining of this is that many operations occur in parallel. A very clear example is when we add two different vectors
  
```{r}
A <- c(2, 4, 6)
B <- c(1, 3, 5)

A + B
```
  the "+" operator acts in *vectorized* way, i.e. it considers all the elements of the vectors and acts with respect to their positions. If we wanted to make this same operation without the vectorization, we should do something as
  
```{r}
z <- vector(length = length(A))
for (i in seq_along(A)) {
  z[i] <- A[i] + B[i]
}
z
```
  
  The actual meaning of this chunk of code will be clear later in the course. The idea here is just to make explicit the simplicity that vectorization implies.
  
  As a side comment, let's mention here that we should avoid the use of **for loops** in R (which does not mean that we are not going to study them!)

## Defining Vectors

There are two three main procedures in R to define a new vector:

 * The **combination function**, c()
 * The **sequences**
 * With the specific **constructors** of each type

### Combination Function

The **c()** function is probably one of the mostly used in any R program. It just takes all the inputs (and these can be as many and in any type you want) and returns the combined vector of them, for example

```{r}
c(1, 2, 3)
```

returns the vector with numbers 1, 2 and 3 as components, while

```{r}
c("me", letters)
```

returns a vector with all the letter in lowercase plus the string "me" at the beginning.

See that the documentation makes explicit the hierarchy that comes implicit in the definition of the funciton. This comes into play when the elements we are combining are of different kind (see later for a deeper explnation on these types), then the function will use

<p style="text-align: center";> 
NULL < raw < logical < integer < double < complex < character < list < expression
</p>

Let's see a few examples

```{r}
c(1, 2L, TRUE)
typeof(c(1, 2L, TRUE))
```

```{r}
c("1", 2L, TRUE)
typeof(c("1", 2L, TRUE))
```

### Sequences

With the combination function we were able to join in one single vector different specific inputs. However, there are many cases in which we need to create a repeated set of values of a given length or a sequence of values with a fixed difference between them (we can think, for example, in a sequence of integer numbers or a sequence of dates).

The easiest way to create a sequence of values is with the colon:

```{r}
1:10
```

this creates a regular sequence of numbers from the first to the second values in the call of the operator with a fixed distance of one between them.

This operator can be used to generate new factors using previous ones via a cross product of them (then again, the actual meaning of the code here will be explain later)

```{r}
F1 <- gl(2,3,10)
F2 <- gl(3,2,10)

F1:F2
```

If we want to speficify a distance bewteen the values other than one we can use the funcion **seq()** to do it. The following function creates a sequence of values from 1 to 5 with jumps of 0.05

```{r}
seq(1, 5, 0.05)
```

If the interval is such that the final value cannot be obtained exactly, the sequence will stop in the number before

```{r}
seq(1, 5, 0.7)
```

There are some important functions related to this seq, in particular we are going to use **seq_along** and **seq.Date** (we recommend you to explore all the others and see if some of them is interesting to you).

The first has already appeared in the for-loop we made before. The idea is that the function creates a sequence with length that of the object in its argument. See the difference in the output of

```{r}
seq_along(F1)
seq_along(length(F1))
```

the first considers the object *F1*, which is a factor of 10 elements and creates a sequence from 1 to 10. The second computes the length of *F1* which is 10, then computes the sequence of this length which is just 1 since the output of the length function is only one number

The other sequence function we are going to consider is one that creates a sequence of **Dates** values

```{r}
seq.Date(as.Date("2019/07/25"), as.Date("2020/07/25"), by = "month")
```

the options that we can use for the **by** are **day**, **week**, **month**, **quarter** or **year**. We can also use our own width in the units of the this *by* argument, then the following will generate the sequence using jumps of three months

```{r}
seq.Date(as.Date("2019/07/25"), as.Date("2020/07/25"), by = "3 month")
```

The last type of sequence we can generate is the **repeated** one:

```{r}
rep(TRUE, 3)
```

with this **rep** function we generate a sequence of the same value repeated the second argument number of times.

### Constructors

In object oriented programming, a **constructor** is a special type of function (subroutine or method) that is used to create a new instance of an object. In the previous methods we have let R decide or impose by hierarchy the type of object that we were going to generate. 

This possibility is not available in many other programming languages and is one of the good a bad things about R: since the interpreter will take care of those details we can go smoothly through the code but sometimes we will find some errors or problems later because we do not have created the type of object we think.

In this sense, working with constructors simplifies things and let us have some control on the objects we are going to use.

We are going to explain the different constructors associated with the different types of atomic vectors which are:

  * Raw
  * Logical
  * Numeric
  * Character
  * Factor

#### Raw

This is the most basic type of data we can find since it treats directly with bytes. One of the cases in which we can find or use this data is when we deal with unstructured datasets or a raw stream of data which has to be converted into structured data.

The constructor of this data is just the **raw** function, then

```{r}
raw(10)
```

this creates an empty raw vector of the specified length. If we want to coerce data into raw we can use the **as.raw** function which returns the hexadecimal value of the input

```{r}
as.raw(124)
```

if we want to turn into raw data a string of characters, we should use the **charToRaw** function

```{r}
charToRaw("124")
```

we can recover the corresponding string with the **rawToChar**

```{r}
rawToChar(as.raw(c(0x31, 0x32, 0x34)))
```

There is another interesting function, the **rawToBits** that returns the bits associated to the raw input.

```{r}
rawToBits(as.raw(124))
```

#### Logical

The next most basic data type is the **logical** one. In this case we will have binary logic data structures with TRUE/FALSE values. The standard constructor is **logical** which receives as input the length of the vector

```{r}
logical(10)
```

Then again, by default it is generated with the equivalent to a sequence of 0's, which in this case are FALSE values.

We can create an particular logical vector with the combination function, such that


```{r}
c(TRUE, FALSE)
```

One particularity of this data type is that it is automatically coerced into numeric 0/1 whenever it is needed by a function, for example

```{r}
mean(c(TRUE, FALSE))
```

In this case, the mean is just returning the proportion of TRUE values in the vector.

We can use the coercion function **as.logical** to turn another data type into logical:

```{r}
as.logical(c(1, 2, 3, -2, 0))
```

see that it returns TRUE whenever there is a value different from 0, and false otherwise.

#### Numeric

There are three types of numbers understood by the interpreter

 * Integer: marked with the **L** flag, these are the usual integer numbers $\mathbb{Z}$
 * Double: these are the default numbers. Equivalent to the real numbers, $\mathbb{R}$
 * Complex: usual complex numbers in $\mathbb{C}$
 
appart from these we can also have hexadecimal numbers.

To generate a numeric vector via a constructor we can use the **vector** function (by default this will generate a logical vector!) which returns a vector of the given length with only 0's as elements

```{r}
vector(mode = "numeric", 10)
```

where **numeric** is synonym to **double**. We can change the **mode** argument to **integer** or **complex** to generate empty vectors of the given length.

```{r}
vector(mode = "complex", 10)
```

this is completely equivalent to using the constructor **complex**

```{r}
complex(10)
```

Just as usual we can use the coercion functions **as.vector**, **as.integer**, **as.numeric** or **as.complex**

#### Character

Character vectors are vectors of literals. Even though these are most commonly generated with the combination function, we can also use the constructor **character** which, again, by default creates an empty vector of characters

```{r}
character(10)
```

See that literals are always defined whenever a values is written inside the quotes, double or single. Then note the difference

```{r}
typeof(c("TRUE", "FALSE", "TRUE"))
typeof(c(TRUE, FALSE, TRUE))
```

the elements of a character vector can be of any lenght and composed by any number of words. The following is a vector of two elements

```{r}
c("hello how", "are you")
```

#### Factor

These are the most particular native data type in R. They represent **categorical variables**. The main two properties of a factor are

  * It is not a numeric vector even if it uses numbers as levels
  * Its levels may or may not contain an implicit order

See that the first property implies that even if the factor describes the answers of a questionnaire as values from 0 to 5, you can never compute quantities associated with numerical variables as could be the mean, the median or the standard deviation.

There are two main constructors to generate a factor:

 * The **gl** (generate levels) function: This function let us create a factor given a nubmer of levels, a vector length and a periodicity of the levels. For example, to generate a factor of with 20 elements and 3 levels each repeated 5 five times

```{r}
gl(3, 5, 20, labels = c("male", "female", "NA"))
``` 
 
 * The **factor** function: This function uses the combination function as input, transforming its values into levels

```{r}
factor(c(1, 0, 0, 1, 1, 0), labels = c("male", "female"))
``` 
 
In setting these factors we have chosen the names of the levels from the very function. However, we may have to choose those level names later in the code. The most straightforward method to change them is with the assignment operator and the combination function. Let's see it
 
```{r}
myfactor <- gl(2, 3, 100)
```

to see the levels of this factor we can do

```{r}
levels(myfactor)
```

Then, suppose we want to use "male" and "female" as the levels

```{r}
levels(myfactor) <- c("male", "female")
```

then now

```{r}
levels(myfactor)
```

later in the course we will see other methods to change this levels.

## Named Vectors

Just as the factors have levels, we can give names to the elements of any vector. This can be done in two different ways

 * With the **setNames** function
 * With the assignment and the **names** function
 
Suppose we generate the following vector

```{r}
myvector <- 1:5
```

let's now generate the vector with the names (this step is not needed but helps with the clarity of the code later)

```{r}
mynames <- c("A", "B", "C", "D", "E")
```

then the assignemnt of the names can be done as

```{r}
myvector <- setNames(myvector, mynames)
```

this is a shortcut of the following assignment

```{r}
names(myvector) <- mynames
```

Now you see in the Global Environment that the vector has a *Named* tag and if you call it, you can see the names. 

Now, in the same way that we may want to assign names, we may want to remove them. This is done with the **unname** function

```{r}
myvector <- unname(myvector)
```

you can see in the Global Environment that the *Named* flag has been removed.

## Subsetting

In many circumstancces we need to retrict the range of a vector and keep only a small subset of its elements. This procedure will be very common in many computations and it is very important that we begin to think about it.

For a one dimensional structure as the vectors we are working with, this susetting is done using the brackets: **vector[condition]** and the condition can be stated calling the element by

 * **position**, or by
 * **name**
 
If we do it by position 

```{r}
myvector[1]
```

if we do it by name


```{r}
myvector["A"]
```

see, that since the name is a character string we must use the quotes to be understood by the interpreter.

We may want to use not just a value, but a **sequence**

```{r}
myvector[1:3]
```

or set of **non sequential** values

```{r}
myvector[c(1, 2, 5)]
```

## Vector Arithmetic

When we talk about vector arithmetic we mean the usual operations we can perform with vectors:

 * sums/substractions
 * products/divisions
 * powers

Remember that we have already mentioned the vectorization nature of all the operations in R. This will have very deep consequences in the output of any operation (the strangest of all in the matrix multiplication and computation of powers)

#### Sum

Consider two vectors

```{r}
A <- 1:4
B <- 4:7
```

the vectorized sum (substruction) is defined as the element-by-element sum

```{r}
A + B
```

Now, what happens if the vectors do not have the same length? R is going to through a warning and will perform the sum following the **recycling rule**

#### The Recycling Rule

This rule implies that the vectors are recycled whenever it is needed, for example

```{r}
A <- 1:4
B <- 4:8
```

now the vector B has five elements while the A has four, then

```{r}
A + B
```

the warning means "hey the vector B is not of length 4, 8, 12 or some other multiple of 4, which is the length of A (in case you did not notice)". But instead of stopping the execution, it just finds the output

|     | 1st | 2nd | 3rd | 4th | 5th   |  
|-----|-----|-----|-----|-----|-------|
| A   | 1   | 2   | 3   | 4   | **1** | 
| B   | 4   | 5   | 6   | 7   | 8     |
| Sum | 5   | 7   | 9   | 11  | **9** |

see that vector A has been recycled once it finished and used from its initial element! The warning just tells that it could not use the whole
vector in the recycling. This implies that if we use vectors of multiple length it will just work silently

```{r}
A <- 1:4
B <- 4:11
```

now the vector B has eight elements while the A has four, then the length of both is multiple of each other, now

```{r}
A + B
```

the sum operator recycles the smaller vector and finds the output without any other problem

|     | 1st | 2nd | 3rd | 4th | 5th | 6th | 7th | 8th |  
|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| A   | 1   | 2   | 3   | 4   | 1   | 2   | 3   | 4   |
| B   | 4   | 5   | 6   | 7   | 8   | 9   | 10  | 11  |
| Sum | 5   | 7   | 9   | 11  | 9   | 11  | 13  | 15  |

#### Product

In R we can perform three different types of products with the vectors

 * A vectorized product done term by term whose result is a vector
 * An inner product whose result is a number
 * An outer product whose result is a matrix

##### Vectorized Product

This is done, like the sum, term by term. Then if

```{r}
A <- 1:4
B <- 4:7
```

the product is

```{r}
A * B
```

|      | 1st | 2nd | 3rd | 4th | 
|------|-----|-----|-----|-----|
| A    | 1   | 2   | 3   | 4   |
| B    | 4   | 5   | 6   | 7   |
| Prod | 4   | 10  | 18  | 28  |

As it is, this product **must not be confused** with the vector multiplication in mathematics! Those other products are the ones we will see in the next two subsections.

We leave for the student checking the outputs of those cases in which the recycling rule comes into play.

##### Scalar Product

The scalar product of two different vectors is defined as as the product of a
(1, n) vector times another (n, 1) vector:

\begin{equation}
\begin{array}{rcl}
A\cdot B & = & (a_1, a_2, a_3)\cdot \begin{pmatrix}b_1\\ b_2\\ b_3\end{pmatrix} \\[2.5ex]
& = & a_1\cdot b_1 + a_2\cdot b_2 + a_3\cdot b_3
\end{array}
\end{equation}

so that the result is just a scalar, i.e. a number or a 1x1 matrix. This is implemented using the binary operator %*%, the for example

```{r}
A %*% A
```

In general, this operator is the usual **matrix multiplication** which will be seen again once we see the matrices and their operations.

##### Outer Product

The %*% operator uses a structure such that it will always try to match the rows and columns to make them conformable. However, if we need the outer product we cannot actually make it that way. In this case we must use the **outer** function as follows

\begin{equation}
\begin{array}{rcl}
A\cdot B & = & \begin{pmatrix} a_1 \\ a_2 \\ a_3 \end{pmatrix} \cdot (b_1, b_2, b_3) \\[2.5ex]
& = & \begin{pmatrix} 
a_1\cdot b_1 & a_1\cdot b_2 & a_1\cdot b_3 \\
a_2\cdot b_1 & a_2\cdot b_2 & a_2\cdot b_3 \\
a_3\cdot b_1 & a_3\cdot b_2 & a_3\cdot b_3 
\end{pmatrix}
\end{array}
\end{equation}

```{r}
outer(A, B)
```
