---
title: "The Apply Family"
subtitle: "PFE I: Session XII"
author: "Prof. Dr. Juanjo Manjarín"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    df_print: kable    
    theme: cerulean
    fig_caption: true
    number_sections: yes
  html_notebook: 
    fig_caption: true
    number_sections: yes
    toc: yes
    toc_depth: 3
  # pdf_document: 
  #   toc: true
---

# The `apply` family

It is commonly accepted that "*nested conditions and loops must be viewed with great suspicion*", as Bjarne Stroustrup put it, and we should add: they are inefficient, then why use them? No actual reason in R when we have the **apply** family: a set of functionals that can (and should) be used as an alternative to the loops. 

Remember that a funtional can be loosely defined as a function of a function, so intuitively, the apply family will have as argument both a data set and a function to be applied to the dataset. There are many different functionals in this family:

 * **apply**: used over arrays
 * **lapply**: used over lists
 * **sapply**: friendly version of *lapply*, but it does not return a consistent output
 * **vapply**: used over vectors
 * **tapply**: used when we want to apply functions grouped by a variable
 * **mapply**: a multivariate version of **sapply**
 * **eapply**: used over environments


## apply

This is the most basic functional of the family and its purpose is to evaluate a function on the boundaries (marginals) of an array. Said so, it is important to keep in mind that we must specify the direction of the marginal in which we evaluate the function, i.e. along rows or columns.

**Create a 4x4 matrix or arbitrary elements and find the mean of its marginals.**

  Let's just create the matrix

```{r}
m <- matrix(seq(1:16), nrow = 4)
```

which, if you remember, creates a matrix with four rows and assigns the inputs filling the columns. Now we use the function **mean** to find the average along the rows and columns:


```{r}
# Along the rows
apply(m, 1, mean)  

# Along the columns
apply(m, 2, mean)
```


so we see that the output of the function is a vector of double elements as the mean of each row or column, and we have used the value of the margin 1  for rows and 2 for columns.
  
## lapply

This function takes as input a **list** and evaluates a function in all the elements of it, returning a list as output.

**Create an inhomogeneous numeric list and find the sum of each variable**

We can create a list as

```{r}
mylist <- list(A = 1:10, B = seq(1,5,0.1), C = seq(1,5,1))
```

we see that in this list al the variables included are numerical and such that *A* has 10 elements running from 1 to 10, *B* has 41 elements running from 1 to 5 in steps of 0.1 and *C* has 5 elements running from 1 to 5 in steps of 1. Thenn it has variables of different lenght, and now we find the sum of each of the variables


```{r}
lapply(mylist, sum)
```

Observe that you cannot use the apply function over a list since it will always return an error due to the dimensions of the list (find this value and check the problem). Can you use lapply to the matrix of the example before? What would be its output?

## sapply

This functional is a user-friendly version of lapply so it can take vectors or lists as arguments, the problem is that it is not consistent and the nature of the output may change from one case to another. This functional evaluates the function on all the elements of the input, then using the two cases above

```{r}
sapply(mylist, sum)
```

In the case of the matrix, the functional considers each element as a different variable in a list, then its sum is extrictly the same value.

```{r}
sapply(m, sum)
```

Now, if we check what type of output we have

```{r}
typeof(sapply(mylist, sum))

typeof(sapply(m, sum))
```

and so, it is not consistent.


## vapply

This functional has some advantages over sapply: first, it always returns a consistent output ans second it is usually a bit faster than sapply. From the syntactic point of view, vapply needs an extra argument for the type of variable and its dimension, then suppose we want to evaluate different properties on a data frame


**Evaluate if the variables in the data frame are numeric and then find the mean of those which are**

```{r}
m <- data.frame(A = 1:10, B = letters[1:10], C = sqrt(seq(1,5.5,0.5)))
```

now we use vapply with a logical output

```{r}
vapply(m, is.numeric, logical(1))
```

then we can compute the mean only for variables A and C, then
  
```{r}
vapply(m[, -2], mean, numeric(1))
```

## tapply

This functional is used whenever we want to apply a function to different subgroups of a data set. Usually these groups come from the values of a categorical variable (factor) and some other will be simply a subgroup determined with a certain condition.

**Suppose you have a data set that consists of the ages of a group of people and their gender, find the average age of each group**

Let's first generate a vector with random ages and then a factor with three levels and of the same lenght as the ages' vector

```{r}
ages <- c(runif(25, 17, 21), runif(21, 18, 22))
gender <- gl(3, 46/3, labels = c("male", "female", "unkown"))
```

then we can use **tapply** as

```{r}
tapply(ages, gender, mean)
```


